---
title: 'Mortality of Alberta: Study of causes of death after experiencing the Great
  Depression.'
author: "Jierui Miao, Shenze Lu"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
date-format: long
abstract: "The recession was horrific, with people facing all sorts of hidden dangers all the time, a collapsing healthcare system, and sky-high drug prices due to inflation. In this report, we choose 2010 as a proxy for the aftermath of the Great Recession and selected the top five causes of death for that year. Using Poisson and negative binomial models, we examine the impact of these causes of death on mortality over the past 20 years and use regression models to predict future trends. The findings suggest that surviving the Great Depression does not mean that chronic diseases, cardiovascular diseases, and tumors have become less harmful to humans. Governments should urge people to be proactive in preventing the onset of disease and raising awareness."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
thanks: "Code and data supporting this analysis is available at: https://github.com/jerry-maker-765/Mortality--Alberta-during-economy-shut.git"
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE}
library(tidyverse)
library(boot)
library(broom.mixed)
library(collapse)
library(dataverse)
library(gutenbergr)
library(janitor)
library(knitr)
library(marginaleffects)
library(modelsummary)
library(rstanarm)
library(ggplot2)
library(lubridate)
library(kableExtra)
library(gridExtra)
library(arrow)
```

\newpage

# Introduction

The dynamics of health status and causes of death in our society changed
as the global economy and healthcare system faced unprecedented
challenges, particularly during the latter part of the Great Depression.
[@cite1] The period not only marked a turning point in the global
financial and economic landscape, but also had a profound impact on all
aspects of society, including public health. Rising unemployment,
falling incomes, and cutbacks in social services during the recession
may have indirectly affected people's health status and lifestyles, thus
having a significant impact on mortality.

This study will focus on the top five causes of death in Alberta in
2010, after the Great Recession, and we find that chronic heart disease
dominated, with malignant neoplasms and acute myocardial infarction also
contributing significantly to mortality during this period. By deeply
analyzing the data from this particular period, we hope to better
understand the specific mechanisms by which economic factors influence
human health and mortality.

In conducting this study, we began by detailing the data sources and
research methods. The dataset contains detailed cause-of-death records,
which provided us with an opportunity to comprehensively analyze the
impact of various mortality factors on population health. In the
modeling section, we built linear regression models with these five
major predictors of cause of death and used negative binomial and
Poisson methods to study their impact on the number of deaths. Through
our analysis, we found that among these five causes, chronic diseases
have the greatest impact on the number of deaths, while Atherosclerotic
cardiovascular disease seems to have the least impact.

We then illustrated the number of deaths caused by these causes each
year from 2001 to 2022 with a line graph showing the change in the
number of deaths caused by each cause. The images show that the number
of deaths due to these diseases has not decreased over the 20-year
period; in fact, the number of deaths due to chronic diseases and
malignant neoplasms is on the rise, which is consistent with the
conclusions we reached in the modeling section.

With this study, we hope to provide insight for policy makers to
emphasize the importance of protecting and promoting public health
during economic downturns and to reduce the negative impact of economic
depression on population health. We also hope to provide an
understanding of the relationship between economic factors and
mortality, and also provide a basis for more robust health policies in
the face of similar economic challenges in the future.

# Data

## Backgrounds

The open dataset used for this analysis is called "Leading Causes of
Death" and is provided by the Government of Alberta. The dataset
provides information on the 30 leading causes of death in Alberta and
ranks these causes of death by total number of deaths. The causes of
death used in the dataset are based on the International Classification
of Diseases, 10th Edition. The dataset includes complete information
from 2001 to 2022 and has been updated annually since 2001, with the
most recent update occurring in September 2023. Thus, 30 leading causes
of death and the number of deaths from those causes have been added to
the dataset each year since 2001, and it is possible for a cause to
appear in the rankings for several consecutive years. The variable "n"
in the dataset will count the number of times each particular cause of
death enters the ranking.

Data cleaning will be performed after loading the data. Since the
dataset contains one NA observation and three (blank) observations in
2014, 2015, and 2018, this may affect the process and results of this
analysis. Therefore, we remove all observations with NA or (blank) from
the dataset before conducting any analysis. The reason we do not use
estimation or any other method to fix missing data is that the missing
data appear in the "Reason" column, which is a "String" column.
Therefore, instead of using estimation or any other method, we will
remove these rows of missing data and there will be no waste of data
since the missing data appears in the "String" column. After that, the
pull function is used to extract only the names of these five causes of
death. Finally, the original dataset, alberta_cod, is updated to retain
those records whose cause-of-death names appear within the top five by
means of the filter function and the %in% operator. After this
processing, the dataset will contain only records related to the top
five causes of death in Alberta in 2010

Our report's original dataset is sourced from the Open Government
program of Alberta province[@Alberta]. And the R packages we used for
programming this report by R language[@citeR] include following
packages:tidyverse[@citetidyverse], boot[@boot],
broom.mixed[@broom_mixed], collapse[@collapse], dataverse[@dataverse],
gutenbergr[@gutenbergr], janitor[@janitor], knitr[@citeknitr],
marginaleffects[@marginaleffects], modelsummary[@modelsummary],
rstanarm[@rstanarm], ggplot2[@rGgplot2], lubridate[@lubridate],
kableExtra[@kableExtra] and gridExtra[@gridExtra].

## Variables
```{r echo=FALSE, message=FALSE}
alberta_cod <- readr::read_csv(here::here("input/data/alberta_cod.csv"))
```

In table 1 the variables of the dataset are introduced.

```{r echo=FALSE, message=FALSE}
#| label: tbl-one
#| tbl-cap: "Types and descriptions for variables."
#| echo: false
#| warning: false
#| message: false

table1 <- data.frame(
  Column = c("Calendar year", "Cause", "Ranking", "Total deaths", "N"),
  Type = c("num", "str", "num", "num", "num"),
  Description = c(
    "A numeric variable. This variable represents the year in which the deaths occurred. It's used to track changes in death rates and causes over time.",
    "A string variable. The cause variable would detail the reason for death, such as specific diseases or accidents. This information is crucial for understanding public health trends and for policy-making.",
    "A numeric variable. Ranking indicate the position of each cause of death in terms of its frequency compared to others within the same year. A lower ranking number(e.g., 1) would imply a higher frequency of deaths due to that cause.",
    "A numeric variable. This variable represents the total number of deaths attributed to each cause in a given year. It's used to measure the impact of different causes of death on the population.",
    "A numeric variable, counts the number of times for a specific cause of death enter the ranking, which is equivalent to the number of times for a specific cause of death appears in the dataset until the latest update."
  )
)


kable(table1, caption = "Types and descriptions for variables.", booktabs = TRUE) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(1, width = "3em") %>% 
  column_spec(2, width = "3em") %>% 
  column_spec(3, width = "33em")
```

# Model

```{r tbl-cap: "Top 10 causes of death in Alberta in 2010.", echo=FALSE, message=FALSE}

alberta_cod |>
  filter(
    calendar_year == 2010,
    ranking <= 10
  ) |>
  mutate(total_deaths = format(total_deaths, big.mark = ",")) |>
  kable(
    caption = "Top 10 causes of death in Alberta in 2010.",
    col.names = c("Year", "Cause", "Ranking", "Deaths", "Years"),
    align = c("l", "r", "r", "r", "r"),
    digits = 0, booktabs = TRUE, linesep = ""
  )
```

In reviewing the processed dataset, the leading causes of death for 2010
were extracted from the dataset, as shown in Table 2.

In this study, we used Poisson and negative binomial models to predict
the total number of deaths caused by a specific cause of death, thus
providing an in-depth analysis of the impact of various causes of death
on the total number of deaths. This study focuses specifically on the
top five causes of death as of 2010: all other forms of chronic ischemic
heart disease; malignant neoplasms of the trachea, bronchus, and lungs;
acute myocardial infarction; organic dementia; and atherosclerotic
cardiovascular disease. By using a count data model, we will explore the
extent to which different causes of death contribute to the number of
deaths and compare which of the Poisson and negative binomial
distributions is more appropriate for describing mortality data in
Alberta. Given the characteristics of count data, the Poisson
distribution, as the simplest regression model, is suitable for cases
where the counts result in non-negative integers. [@poisson] However,
considering the over-dispersion that actual data may exhibit, the
negative binomial model, which is an extension of the Poisson model to
include a stochastic component that reflects uncertainty in event rates,
[@negatvie] may be more appropriate for use in the analysis of the data
in this study. With this comparison, this study aims to provide a more
accurate model to predict the impact of different causes of death on the
total number of deaths, thus providing a scientific basis for the
formulation of targeted public health policies.

```{r echo=FALSE, message=FALSE, warning=FALSE}


alberta_cod_top_five <-
  alberta_cod |>
  filter(
    calendar_year == 2010,
    
  ) |>
  slice_max(order_by = desc(ranking), n = 5) |>
  pull(cause)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
alberta_cod <-
  alberta_cod |>
  filter(cause %in% alberta_cod_top_five)
```

\newpage

## Poisson model



```{r tbl-cap: "Comparison of Poisson and negative binomial results.", echo=FALSE, message=FALSE, warning=FALSE}
library(rstanarm)
library(knitr)
library(modelsummary)

cause_of_death_alberta_poisson <-
  stan_glm(
    total_deaths ~ cause,
    data = alberta_cod,
    family = poisson(link = "log"),
    seed = 800,
    refresh = 0  # Suppresses the Stan sampler output
  )

cause_of_death_alberta_neg_binomial <-
  stan_glm(
    total_deaths ~ cause,
    data = alberta_cod,
    family = neg_binomial_2(link = "log"),
    seed = 800,
    refresh = 0  # Suppresses the Stan sampler output
  )

modelsummary(
  list(
    "Poisson" = cause_of_death_alberta_poisson,
    "Negative binomial" = cause_of_death_alberta_neg_binomial
  ),
  title = "Comparison of Poisson and negative binomial results" 
)
```

$$ log(E(y)) = 7.038 + 0.446 x_{1}- 0.436 x_{2} + 0.223 x_{3} + 0.045 x_{4}$$

{#eq-equation1}

The Poisson model analyzes the top five causes of death in 2010 and how
they affect the expected total number of deaths, denoted as E(y).
The intercept corresponds to the impact of Acute Myocardial Infarction.
Variables x1 to x4 represent the effects of four different
causes of death, with their coefficients indicating the direction of
impact (positive values indicate a positive effect, negative values
indicate a negative effect): - x1 represents the impact of "All
other forms of chronic ischemic heart disease." - x2 represents
the impact of "Atherosclerotic cardiovascular disease," with a negative
coefficient, indicating that this cause is negatively correlated with
the number of deaths. - x3 represents the impact of "Malignant
neoplasms of trachea, bronchus, and lung." - x4 represents the
impact of "Organic dementia."

This model indicates that x1 (other forms of chronic ischemic
heart disease) has a positive impact on the number of deaths, meaning it
increases the total number of deaths. On the other hand, x2
(atherosclerotic cardiovascular disease) has a negative impact on the
number of deaths, implying that it reduces the total number of deaths.
The positive coefficients for x3 and x4 demonstrate that
they also contribute to an increase in the total number of deaths,
although the magnitude of their impact varies.

$$ E(y) = e^{7.038 + 0.446 x_{1} - 0.436 x_{2} + 0.223 x_{3} + 0.045 x_{4}}$$
{#eq-equation2}

This equation is derived from the exponential transformation of the
Poisson regression model, providing a more intuitive way to understand
how various factors affect the expected value of the number of deaths,
denoted as E(y). For instance, if the value of a factor increases
and its corresponding coefficient is positive, then E(y), the
expected total number of deaths, will increase. Conversely, if a
factor's coefficient is negative, then E(y) will decrease. Such a
model allows us to quantify the impact of specific health conditions on
the total number of deaths, offering a scientific basis for the
formulation of public health strategies and the allocation of resources.

## negative binominal model

Considering the possibility of overdispersion after completing the
Poisson model analysis, we decided to conduct an analysis using the
negative binomial regression model next. This approach is often taken
when the variance in the data significantly exceeds the mean, which can
happen in count data like the number of deaths. The negative binomial
regression can handle overdispersion by introducing an extra parameter
to account for the variability, making it a suitable alternative for
more accurately modeling the data under such conditions.

$$log(E(y)) = 7.04 + 0.444 x_{1} - 0.437 x_{2} + 0.222 x_{3} + 0.046 x_{4}$$
{#eq-equation3}

$$ E(y) = e^{7.04 + 0.444 x_{1} - 0.437 x_{2} + 0.222 x_{3} + 0.046 x_{4}}$$
{#eq-equation4}

## Poisson vs Negative Binomial model

```{r fig.cap="Comparing posterior prediction checks for Poisson and negative binomial models", echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=3.2}
library(rstanarm)
library(bayesplot)



poisson_plot <- pp_check(cause_of_death_alberta_poisson) +
  theme(legend.position = "bottom") +
  ggtitle("Poisson Model Check for Alberta Cause of Death")

neg_binomial_plot <- pp_check(cause_of_death_alberta_neg_binomial) +
  theme(legend.position = "bottom") +
  ggtitle("Negative Binomial Model Check for Alberta Cause of Death")

poisson_plot
neg_binomial_plot
```

We created graphs for both the Poisson model and the negative binomial
model to observe if these models are suitable for estimation.

The first graph shows a distribution with clear peaks and troughs. The
predictions generated by the model align well with the actual data in
some areas but show significant deviations in others. This discrepancy
might indicate that the model has not captured all the variations in the
data, especially at the peaks and troughs.

The second graph presents a relatively smooth and continuous
distribution, where the model's predictions (in light blue) closely
follow the actual observations (in dark blue). This typically indicates
that the model's predictions are more consistent with the actual data.

It is evident that the precision of the Poisson model's predictions is
not as high as that of the negative binomial model. To further confirm
our observation, we can use the resampling method Leave-One-Out (LOO)
Cross-Validation (CV) to compare the models.

```{r tbl-cap: "ELPD and SE difference in Negative Binomial and Poisson models.", echo=FALSE, message=FALSE, warning=FALSE}

poisson <- loo(cause_of_death_alberta_poisson, cores = 2)
neg_binomial <- loo(cause_of_death_alberta_neg_binomial, cores = 2)

compare <- loo_compare(poisson, neg_binomial)


compare_df <- as.data.frame(compare)

rownames(compare_df) <- c("Alberta CoD under Neg. Binomial", "Alberta CoD under Poisson")
colnames(compare_df)[colnames(compare_df) == "elpd_diff"] <- "ELPD difference"
colnames(compare_df)[colnames(compare_df) == "se_diff"] <- "SE difference"

kable(
  compare_df[, c("ELPD difference", "SE difference")],
  caption = "ELPD and SE difference in Negative Binomial and Poisson models."
) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

In the LOO comparison results, the LOOIC difference for the cause_of_death_alberta_neg_binomial model is 0.0, while for the
cause_of_death_alberta_poisson model, it is -4954.0. This negative value
indicates that the predictive performance of the Poisson model is
significantly worse than that of the negative binomial model. The second
column shows the change in the effective sample size (displayed here as
1208.0), which is an estimate of the uncertainty in the model
comparison.

These results indicate that among the two models, the negative binomial
regression model (cause_of_death_alberta_neg_binomial) performs better
under Leave-One-Out cross-validation, meaning it has better predictive
performance for the data. This further supports the previous observation
that the negative binomial model is more suitable for handling
overdispersed count data.

# result

```{r  fig.width=10, fig.height=8, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(RColorBrewer)

# Adjusting the plot dimensions can make it more readable
alberta_cod %>%
  ggplot(aes(x = calendar_year, y = total_deaths, color = cause)) +
  geom_line(size = 1.2) +  # Thicker lines for better visibility
  theme_minimal(base_size = 14) +  # Larger base font size
  scale_color_brewer(palette = "Set1") +
  labs(
    x = "Year", 
    y = "Annual number of deaths in Alberta", 
    title = "Figure 2: Annual number of deaths in Alberta for the top-five causes in 2010, since 2001."
  ) +
  facet_wrap(vars(cause), dir = "v", ncol = 1, scales = "free_y") +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis text if needed
    axis.text.y = element_text(size = 12),  # Adjust y-axis text size
    strip.text = element_text(size = 12),  # Adjust facet label size
    plot.title = element_text(size = 16, face = "bold")  # Adjust title size and make it bold
  )
```

Using line graphs, we looked at trends in the number of deaths from
these five causes of death from 2001 to 2022. In Alberta, trends over
time in the annual number of deaths from the five leading causes of
death show different patterns. Annual deaths from acute myocardial
infarction and all other forms of chronic ischemic heart disease have
remained relatively stable, with no significant upward or downward
trends. In contrast, the number of deaths from atherosclerotic
cardiovascular disease has declined slightly. On the other hand, the
number of deaths from malignant tumors of the trachea, bronchus and
lungs rose slightly, while the number of deaths from organic dementia
showed a more significant increasing trend. Such changes may reflect the
aging of the population and changes in the health care field.

```{r echo=FALSE, message=FALSE, warning=FALSE}

name <- c(" chronic ischemic heart disease", "Atherosclerotic cardiovascular disease", "Malignant neoplasms of trachea, bronchus and lungr", "Organic dementia", "Acute myocardial infarction")

expected_deaths_poisson <- rep(exp(7.038), length(name)) 
expected_deaths_negbin <- rep(exp(7.04), length(name))

expected_deaths_poisson[1] <- exp(7.038 + 0.446)  
expected_deaths_poisson[2] <- exp(7.038 - 0.436)  
expected_deaths_poisson[3] <- exp(7.038 + 0.223)  
expected_deaths_poisson[4] <- exp(7.038 + 0.045)  


expected_deaths_negbin[1] <- exp(7.04 + 0.444)  
expected_deaths_negbin[2] <- exp(7.04 - 0.437)  
expected_deaths_negbin[3] <- exp(7.04 + 0.222)  
expected_deaths_negbin[4] <- exp(7.04 +0.046)  


estimate <- data.frame(
  Cause = name,
  Poisson = expected_deaths_poisson,
  `Negative Binomial` = expected_deaths_negbin
)

# Creating the markdown table with kable and adding a caption
kable(estimate, format = "markdown", caption = "Comparison of Poisson and negative binomial estimates") %>%
  kable_styling("striped", full_width = F)
```

Through the models we have developed, we can simulate the number of
deaths and ultimately make a comparison. The comparison shows that the
negative binomial model's results are closer to the actual situation for
deaths caused by "Malignant neoplasms of trachea, bronchus, and lung"
(1425 vs. 1431). The Poisson model's predictions for deaths caused by
"Organic dementia" and "Atherosclerotic cardiovascular disease" are
closer to the real situation.

This conclusion might seem to conflict with the earlier observation that
the negative binomial model fits the real situation more closely.
However, this does not imply that the negative binomial model is less
accurate. Earlier studies have shown that the negative binomial model
has better predictive power. But under certain conditions (i.e., when
the data is not overly dispersed), the Poisson model might be
sufficient. In most real-world situations, especially when data exhibits
overdispersion, the negative binomial model often provides more accurate
predictions. The negative binomial model is better suited to adapt to
different data fluctuations. This might explain why the Poisson model
appears to perform better in simulating data in this context.

# Discussion

From the analysis of trends in mortality rates in Alberta from 2000 to
2020 reveals distinct patterns in public health outcomes. Specifically,
the data demonstrates a stable decreasing trend in deaths attributed to
cardiovascular diseases, which could be caused by advancements in
medical technology, enhanced emergency response protocols, and improved
management of risk factors such as hypertension and high cholesterol.
Public education on heart health appears to have contributed positively
to these outcomes. In contrast, a marked increase in mortality rates due
to organic dementia points towards demographic shifts towards an older
population and advancements in diagnostic techniques. Additionally, an
uptick in deaths from malignant respiratory system neoplasms underscores
the potential impacts of long-standing smoking habits and environmental
factors such as industrial pollution, coupled with gaps in screening
processes. These findings highlight the successes in combating
cardiovascular diseases through innovation and public health
initiatives, while also stressing the need for focused efforts on
dementia and certain cancers, with an emphasis on prevention, early
detection, and novel treatment approaches.

what is more,our study examined the effectiveness of Poisson and
Negative Binomial Regression models in estimating the expected total
deaths (E(y)) from the top five causes of death before 2010. Through
thorough model evaluation, including graphical analysis and
Leave-One-Out Cross-Validation (LOO CV), our findings highlight
significant differences in model performance and applicability.

The Poisson Regression Model is known for its simplicity and the
assumption that the mean and variance are equal, providing a basic
understanding of how various causes of death impact expected mortality
numbers. However, this model shows limitations in capturing the
complexity of the data, especially in cases of overdispersion (where
variance exceeds the mean). This is evident from the observed
differences between model predictions and actual data, particularly at
points of fluctuation like peaks and troughs.

In contrast, the Negative Binomial Regression Model offers a more robust
alternative by incorporating an additional parameter to account for
variance, addressing the challenge of overdispersion. Comparative
analysis shows that the Negative Binomial model has superior predictive
performance, with predictions more consistent with actual observations
and significantly better LOO CV scores. The effectiveness of this model
is demonstrated by its ability to accurately simulate the number of
deaths from specific causes, such as "malignant neoplasms of the
trachea, bronchus, and lung," showing its adaptability to data
fluctuations.

Also, while the Poisson model produced closer predictions in some
instances, such as deaths caused by "organic dementias" and
"atherosclerotic cardiovascular diseases," this did not detract from the
overall superior performance of the Negative Binomial model in handling
overdispersed count data. Our results emphasize the importance of model
selection in epidemiological research, advocating for the use of the
Negative Binomial model in cases of overdispersion to ensure more
accurate and reliable estimation of mortality numbers.

This comparative analysis underscores the urgent need for researchers to
carefully consider the characteristics of their data and the fundamental
assumptions of their statistical models. By doing so, we can enhance the
accuracy of public health predictions and interventions, ultimately
contributing to more effective health resource management and better
health outcomes.

\newpage

# references
